---
title: "Cluster randomized trial sample size function"
subtitle: "means"
author: "Zijun Liu and Ben Arnold"
date: 'updated: `r Sys.time()`'
output:
  html_document:
    highlight: haddock
    theme: default
    code_folding: show
    df_print: paged
    toc: yes
    toc_depth: '3'
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

# preamble
```{r preamble}
# load packages 
# (only required for the example) 
library(tidyverse)

```

# ss_cluster_mean

Estimate the number of clusters requred per treatment arm, $c$, for a parallel cluster randomized trial with equal allocation. This is equation 7.12 (pg 135) in Hayes and Moulton (2017) Cluster randomised trials. (Chapman and Hall/CRC).

\begin{equation}
c = 1 + \left( z_{\alpha/2} + z_\beta \right)^2 \frac{(\sigma_0^2 + \sigma_1^2) \times \left[ 1 + (m-1)\rho \right]}{m(\mu_0 - \mu_1)^2}
\end{equation}

where:

* $c$ : is the number of clusters _per arm_
* $\alpha$ : is the Type-I error rate (divided by 2 for a two-sided test)
* $\beta$ : is the Type-II error rate (1-$\beta$ is power)
* $\mu_0$, $\mu_1$ : are the _true_ means in the control and intervention groups
* $\sigma_0$, $\sigma_1$: are the standard deviations in the control and intervention groups
* $m$ : is the average number of individuals per cluster
* $\rho$ : is the intra-cluster correlation coefficient (ICC) for repeated measures within cluster

```{r ss_cluster_mean}
#----------------------------
# ss_cluster_mean
#
# calculate the number of clusters required per arm 
# for defined power (1-beta), two-sided alpha, and
# assumed means in the two groups. Note, the detectable
# effect is mu1 - mu0
#
# This is equation 7.12 in Hayes & Moulten 2017. Cluster randomised trials.
#
# @param alpha : type I error, e.g. 0.05
# @param beta  : type II error, e.g. 0.2, (1-beta = power)
# @param mu0   : outcome mean in the control group
# @param mu1   : outcome mean in the intervention group
# @param sigma0: outcome stnadard deviation in the control group
# @param sigma1: outcome standard deviation in the intervention group
# @param m     : average number of measurements per cluster
# @param rho   : intra-cluster correlation coefficient (ICC) for the outcome
#
# returns: 
# @numeric c : number of clusters per arm (not rounded / not integer)
ss_cluster_mean <- function(alpha, beta, sigma0, sigma1, m, rho, mu0, mu1) {
    zab <- (qnorm(1-alpha/2) + qnorm(1-beta))^2
    num1 <- sigma0^2 + sigma1^2
    num2 <- 1 +(m-1)*rho
    denom <- m * (mu0-mu1)^2
    c <- 1 + zab * (num1*num2)/denom
    c
}

```

# example

For a range of effect sizes, calculate the required number of clusters.

```{r ss_cluster_mean example}
# assume mean in control is 1, sd in both arms is 0.25
# vary the mean in the intervention arm from 1.3 down to 1.1
# assume clusters include 30 observations, with ICC = 0.05
# assume a 2-sided type I alpha of 0.05, and 90% power (beta = 0.1)
mu1s <- seq(1.1, 1.3, by=0.05)
cs <- map_dbl(mu1s, ss_cluster_mean, alpha=0.05, beta=0.1, sigma0 = 0.25, sigma1 = 0.25, m=30, rho=0.05, mu0=1)

```

Plot results

```{r ss_cluster_mean example figure}
d_plot <- data.frame(mu1=mu1s, delta = mu1s-1, c=cs)
ggplot(data = d_plot, aes(x = delta, y = c)) +
  geom_point() +
  geom_line() +
  geom_text(aes(label = paste0(sprintf("%1.0f",c)) ), nudge_x = 0.002, nudge_y = 0.5) +
  labs(x = expression(paste("effect size (", mu[0] - mu[1],")")), y = "clusters per arm (c)") +
  theme_minimal()
```
  

# session info
```{r session info}
sessionInfo()
```
