---
title: "Cluster randomized trial sample size function"
subtitle: "proportions"
author: "Zijun Liu and Ben Arnold"
date: 'updated: `r Sys.time()`'
output:
  html_document:
    highlight: haddock
    theme: default
    code_folding: show
    df_print: paged
    toc: yes
    toc_depth: '3'
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

# preamble
```{r preamble}
# load packages 
# (only required for the example) 
library(tidyverse)

```

# ss_cluster_prop

Estimate the number of clusters requred per treatment arm, $c$, for a parallel cluster randomized trial with equal allocation. This is equation 7.11 (pg 135) in Hayes and Moulton (2017) Cluster randomised trials. (Chapman and Hall/CRC).

\begin{equation}
c = 1 + \left( Z_{\alpha/2} + Z_\beta \right)^2 \frac{\left[ \pi_0(1-\pi_0) + \pi_1(1-\pi_1) \right] \times \left[ 1 + (m-1)\rho \right]}{m(\pi_0 - \pi_1)^2}
\end{equation}

where:

* $c$ : is the number of clusters _per arm_
* $\alpha$ : is the Type-I error rate (divided by 2 for a two-sided test)
* $\beta$ : is the Type-II error rate (1-$\beta$ is power)
* $\pi_0$, $\pi_1$ : are the proportions in the control and intervention groups
* $m$ : is the average number of individuals per cluster
* $\rho$ : is the intra-cluster correlation coefficient (ICC) for repeated measures within cluster

```{r ss_cluster_prop}
#----------------------------
# ss_cluster_prop
#
# calculate the number of clusters required per arm 
# for defined power (1-beta), two-sided alpha, and
# assumed prevalence in the two groups. Note, the detectable
# effect is pi1 - pi0
#
# This is equation 7.11 in Hayes & Moulten 2017. Cluster randomised trials.
#
# @param alpha : type I error, e.g. 0.05
# @param beta  : type II error, e.g. 0.2, (1-beta = power)
# @param pi0   : outcome proportion in the control group
# @param pi1   : outcome proportion in the intervention group
# @param m     : average number of measurements per cluster
# @param rho   : intra-cluster correlation coefficient (ICC) for the outcome
#
# returns: 
# @numeric c : number of clusters per arm (not rounded / not integer)
ss_cluster_prop <- function(alpha, beta, pi0, pi1, m, rho) {
    zab <- (qnorm(1-alpha/2) + qnorm(1-beta))^2
    num1 <- pi*(1-pi0) + pi1*(1-pi1)
    num2 <- 1 + (m-1)*rho
    denom <- m * (pi0-pi1)^2
    c <- 1 + zab * (num1*num2)/denom
    c
}

```

# example

For a range of effect sizes, calculate the required number of clusters.

```{r ss_cluster_prop example}
# assume prevalence in control is 20%
# vary the proportion in the intervention arm from 18% down to 10%
# assume clusters include 30 observations, with ICC = 0.05
# assume a 2-sided type I alpha of 0.05, and 90% power (beta = 0.1)
pi1s <- seq(0.10,0.18,by=0.01)
cs <- map_dbl(pi1s, ss_cluster_prop, alpha=0.05, beta=0.2, pi0= 0.2, m=30, rho=0.05)

```

Plot results

```{r ss_cluster_prop example figure}
d_plot <- data.frame(pi1=pi1s, delta = 0.2-pi1s, c=cs)
ggplot(data = d_plot, aes(x = delta, y = c)) +
  geom_point() +
  geom_line() +
  geom_text(aes(label = paste0(sprintf("%1.0f",c)) ), nudge_x = 0.002, nudge_y = 200) +
  labs(x = expression(paste("effect size (", pi[1] - pi[0],")")), y = "clusters per arm (c)") +
  theme_minimal()
  

```

# session info
```{r session info}
sessionInfo()
```